\pkg{sdmTMB}: An \proglang{R} Package for Fast, Flexible, and User-Friendly Generalized Linear Mixed Effects Models with Spatial and Spatiotemporal Random Fields

### 2024-06-01 doc/paper-jss/sdmTMB-paper.Rnw
Edits

\added{
Users willing to replace the default \R\ \pkg{BLAS} \citep{blas}
% and \pkg{LAPACK} \citep{lapack}
library with an optimized version \citep[e.g., \pkg{openBLAS};][]{openblas}
can expect up to an order of magnitude increase in model fitting speed.
Suggestions are included in the package README file.
}

@@

\added{
We can then calculate any performance metric of interest for comparison.
\Rev{A2}A common metric would be the log score or log predictive (likelihood) density (lpd) of the left-out data \citep[][]{geisser1979, vehtari2017}
}

\added{
$$
\mathrm{lpd} = \sum_{i=1}^{n} \ell (\widehat{y_i} | y_i),
$$
}

\added{
where $n$ represents the number of left-out data points, $\ell$ represents the log likelihood, $y_i$ represents left-out data point $i$, and $\widehat{y_i}$ represents the prediction for left-out data point $i$ with the fixed effects at their MLEs and random effects at their empirical Bayes estimates.
}

@@

Our model can be written as
\[
\begin{aligned}
\mathbb{E}[y_{\bm{s},t}] &= \mu_{\bm{s},t},\\
\mu_{\bm{s},t} &=
\exp \left( \bm{X}^{\mathrm{main}}_{\bm{s},t} \bm{\beta} +
O_{\bm{s},t} +
\bm{X}^{\mathrm{tvc}}_{\bm{s},t} \gamma_t +
\omega_{\bm{s}} +
\epsilon_{\bm{s},t}
\right),\\
\bm{\omega} &\sim \operatorname{MVNormal} \left( \bm{0}, \bm{\Sigma}_\omega \right),\\
\bm{\epsilon}_{t} &\sim \operatorname{MVNormal} \left( \bm{0}, \bm{\Sigma}_{\epsilon} \right),
\end{aligned}
\]
where \(\bm{\beta}\) are coefficients associated with the main effects, \(O_{\bm{s},t}\) represents the offset (here, log area swept), \(\gamma_t\) represents the time-varying coefficients, \(\omega_{\bm{s}}\) is the value from a spatial field (representing constant latent spatial effects) and \(\epsilon_{\bm{s},t}\) is a value from a spatiotemporal field  (representing latent spatial effects that vary by year).
The temporally varying intercepts \(\gamma_t\) are modeled as a stationary AR(1) process,
\[
\begin{aligned}
  \gamma_{t=1} &\sim \operatorname{Normal} \left(0, \sigma^2_{\gamma} \right),\\
  \gamma_{t>1} &\sim \operatorname{Normal} \left(\rho_\gamma\gamma_{t-1}, \sqrt{1 - \rho_\gamma^2} \sigma^2_{\gamma} \right),
\end{aligned}
\]
where \(\rho_\gamma\) represents the correlation between intercepts \(\gamma\) at time $t-1$ and $t$ and $\sigma^2_{\gamma}$ represents the marginal variance of this process.
We include the argument \code{extra\_time}, which represents all time steps to include in the latent process (and all time steps for which we may wish to predict), such that autoregressive processes are applied to equally spaced annual time steps.

@@

@@

We ran a simple speed comparison between \sdmTMB, \INLA/\pkg{inlabru}, and \pkg{mgcv} for fitting an SPDE spatial random field model to 1,000, \added{10,000, or 100,000 data points} with Gaussian error across a range of mesh resolutions (Figure~\ref{fig:timing}, Appendix \ref{app:speed}).
Our test was restricted to one core and default \R\ algebra libraries.
% \sdmTMB\ (and due to having the same \TMB\ backend, \pkg{VAST}) were fastest with a spatial model up to at least \textasciitilde 1000 mesh vertices at approximately a seven-times (at 1165 vertices) to 30-times (at 125 vertices) speed increase over \INLA/\pkg{inlabru}.
% These speed increases can allow for more rapid and thorough model exploration and experimentation with a class of computationally intensive models.
\added{
With up to 10,000 rows of data, \sdmTMB\ was fastest at approximately a three to 13-fold speed increase over \INLA/\pkg{inlabru} (Figure~\ref{fig:timing}a--b).
At larger sample sizes, \pkg{inlabru} was more affected by mesh resolution than \sdmTMB\ (Figure~\ref{fig:timing}c).
\pkg{mgcv} was most affected by mesh resolution (Figure~\ref{fig:timing})---timing was on par or faster than the other packages at low ($\sim$ 200) mesh resolutions but much slower as mesh complexity grew into commonly used ranges.
\pkg{spaMM} scaled well to large datasets with high mesh resolutions (Figure~\ref{fig:timing}c), but can only fit spatial and not spatiotemporal models.
At low mesh resolutions, a large portion of the differences in timing are likely a result of initial model processing and not estimation itself.
}
Speed increases can allow for more rapid and thorough model exploration and experimentation with a class of computationally intensive models.
For users ultimately interested in Bayesian inference, the approximate Bayesian inference offered by \INLA/\pkg{inlabru} is likely to be considerably faster than passing the same model from \sdmTMB/\pkg{VAST} to \pkg{tmbstan} for full MCMC-based Bayesian inference.

\begin{figure}[htb]
\centering
% \includegraphics[width=3.5in]{../../figs/timing-spatial-2023-03-16-xkcd.pdf}
\includegraphics[width=\textwidth]{figs/timing-spatial-multipanel.pdf}
\caption[]{
Comparison of time to fit an SPDE spatial random field model with an intercept, one fixed-effect predictor, Gaussian error, and a sequence of SPDE resolutions to three dataset sizes.
Lines represent means and ribbons 90\% quantiles across 50 random iterations.
\pkg{VAST} and \pkg{tinyVAST} should be similar to \sdmTMB\ and so are not shown.
\pkg{inlabru} (version 2.10.1) used the empirical Bayes integration strategy and Gaussian approximation with \code{bru\_max\_iter = 1}, and the \fct{like} formulation.
\pkg{mgcv} (version 1.9.1) used \fct{bam}, \code{method = `fREML'}, and discretized covariates \citep{miller2019}.
Note that \pkg{spaMM} (version 4.4.16) only fits spatial, not spatiotemporal, models.
All platforms were restricted to one core and could be faster with parallel computation or optimized algebra libraries.
See Figure~\ref{fig:timing2} for a version with log-log axes.
}
\label{fig:timing}
\end{figure}

@@

We conducted 50 simulation iterations per model and mesh resolution and show timing results for the mean, 10\%, and 90\% quantile values.
The models were fit on a 2023 16 inch M2 MacBook Pro with an Apple M2 Pro 10-core CPU and 32 GB of RAM in R 4.2.2 and the default \pkg{BLAS} \citep{blas} library packaged with \R \citep{r2024}.

>>>>

\added{
Users willing to replace the default \R\ \pkg{BLAS} (Basic Linear Algebra Subprograms) \citep{blas}
% and \pkg{LAPACK} \citep{lapack}
library with an optimized version \citep[e.g., \pkg{openBLAS};][]{openblas}
can expect up to an order of magnitude increase in model fitting speed for complex models.
Suggestions are included in the package README file.
}

@@

\added{
We can then calculate any performance metric of interest for comparison.
\Rev{A2}A common metric would be the log score or log predictive (likelihood) density (lpd) of the left-out data \citep[][]{geisser1979, vehtari2017},
}

\added{
$$
\mathrm{lpd} = \sum_{i=1}^{n} \log \mathcal{L} (\widehat{y_i} | y_i),
$$
}

\added{
where $n$ represents the number of left-out data points, $\log \mathcal{L}$ represents the log likelihood, $y_i$ represents left-out data point $i$, and $\widehat{y_i}$ represents the prediction for left-out data point $i$ with the fixed effects at their MLEs and random effects at their empirical Bayes estimates.
}

@@

Our model can be written as
\[
\begin{aligned}
\mathbb{E}[y_{\bm{s},t}] &= \mu_{\bm{s},t},\\
\mu_{\bm{s},t} &=
\exp \left( \bm{X}^{\mathrm{main}}_{\bm{s},t} \bm{\beta} +
O_{\bm{s},t} +
\bm{X}^{\mathrm{tvc}}_{\bm{s},t} \gamma_t +
\omega_{\bm{s}} +
\epsilon_{\bm{s},t}
\right),\\
\bm{\omega} &\sim \operatorname{MVNormal} \left( \bm{0}, \bm{\Sigma}_\omega \right),\\
\bm{\epsilon}_{t} &\sim \operatorname{MVNormal} \left( \bm{0}, \bm{\Sigma}_{\epsilon} \right),
\end{aligned}
\]
where \(\bm{\beta}\) are coefficients associated with the main effects, \(O_{\bm{s},t}\) represents the offset (here, log area swept), \(\gamma_t\) represents the time-varying coefficients, \(\omega_{\bm{s}}\) is the value from a spatial field (representing constant latent spatial effects) and \(\epsilon_{\bm{s},t}\) is a value from a spatiotemporal field  (representing latent spatial effects that vary by year).
The temporally varying intercepts \(\gamma_t\) are modeled as a stationary AR(1) process,
\[
\begin{aligned}
  \gamma_{t=1} &\sim \operatorname{Normal} \left(0, \sigma^2_{\gamma} \right),\\
  \gamma_{t>1} &\sim \operatorname{Normal} \left(\rho_\gamma\gamma_{t-1}, \sqrt{1 - \rho_\gamma^2} \sigma^2_{\gamma} \right),
\end{aligned}
\]
where \(\rho_\gamma\) represents the correlation between intercepts \(\gamma\) at time $t-1$ and $t$ and $\sigma^2_{\gamma}$ represents the marginal variance of this process.
\added{We include the argument \code{extra\_time}, which represents all time steps to include in the latent process (and all time steps for which we may wish to predict), such that autoregressive processes are applied to equally spaced annual time steps.}

@@

\added{
As an example of estimation time for these complex spatiotemporal models, the Tweedie model (\code{fit\_tw}), Poisson-link delta-lognormal model (\code{fit\_dpl}), and Poisson-link delta-lognormal model with autogressive random fields (\code{fit\_dpl\_ar1}) took approximately 10, 30, and 90 seconds to fit, respectfully, on an Apple MacBook Pro with an M2 Pro processor and with Apple's \pkg{vecLib} implementation of \pkg{BLAS} in \R\ 4.4.0.
}

@@

We ran a simple speed comparison between \sdmTMB, \INLA/\pkg{inlabru}, and \pkg{mgcv} for fitting an SPDE spatial random field model to 1,000, \added{10,000, or 100,000 data points} with Gaussian error across a range of mesh resolutions (Figure~\ref{fig:timing}, Appendix \ref{app:speed}).
Our test was restricted to one core and default \R\ algebra libraries.
% \sdmTMB\ (and due to having the same \TMB\ backend, \pkg{VAST}) were fastest with a spatial model up to at least \textasciitilde 1000 mesh vertices at approximately a seven-times (at 1165 vertices) to 30-times (at 125 vertices) speed increase over \INLA/\pkg{inlabru}.
% These speed increases can allow for more rapid and thorough model exploration and experimentation with a class of computationally intensive models.
\added{
With up to 10,000 rows of data, \sdmTMB\ was fastest at approximately a three- to 13-fold speed increase over \INLA/\pkg{inlabru} (Figure~\ref{fig:timing}a--b).
At larger sample sizes, \pkg{inlabru} was more affected by mesh resolution than \sdmTMB\ (Figure~\ref{fig:timing}c).
\pkg{mgcv} was most affected by mesh resolution (Figure~\ref{fig:timing})---timing was on par or faster than the other packages at low ($\sim$ 200) mesh resolutions but much slower as mesh complexity grew into commonly used ranges.
\pkg{spaMM} scaled well to large datasets with high mesh resolutions (Figure~\ref{fig:timing}c), but can only fit spatial GMRFs.
At low mesh resolutions, a large portion of the differences in timing are likely a result of initial model processing and not estimation itself.
}
Speed increases can allow for more rapid and thorough model exploration and experimentation with a class of computationally intensive models.
For users ultimately interested in Bayesian inference, the approximate Bayesian inference offered by \INLA/\pkg{inlabru} is likely to be considerably faster than passing the same model from \sdmTMB/\pkg{VAST} to \pkg{tmbstan} for full MCMC-based Bayesian inference.

\begin{figure}[htb]
\centering
% \includegraphics[width=3.5in]{../../figs/timing-spatial-2023-03-16-xkcd.pdf}
\includegraphics[width=\textwidth]{figs/timing-spatial-multipanel.pdf}
\caption[]{
Comparison of time to fit an SPDE spatial random field model with an intercept, one fixed-effect predictor, Gaussian error, and a sequence of SPDE resolutions to \added{three dataset sizes}.
Lines represent means and ribbons 90\% quantiles across 50 random iterations.
\pkg{VAST} and \pkg{tinyVAST} should be similar to \sdmTMB\ and so are not shown.
\pkg{inlabru} \added{(version 2.10.1)} used the empirical Bayes integration strategy and Gaussian approximation with \code{bru\_max\_iter = 1}, and the \fct{like} formulation.
\pkg{mgcv} \added{(version 1.9.1)} used \fct{bam}, \code{method = `fREML'}, and discretized covariates \citep{miller2019}.
Note that \pkg{spaMM} \added{(version 4.4.16)} only fits spatial, not spatiotemporal, models.
All platforms were restricted to one core and could be faster with parallel computation or optimized algebra libraries.
\added{See Figure~\ref{fig:timing2} for a version with log-log axes.}
}
\label{fig:timing}
\end{figure}

@@

We conducted 50 simulation iterations per model and mesh resolution and show timing results for the mean, 10\%, and 90\% quantile values.
The models were fit on a 2023 16 inch M2 MacBook Pro with an Apple M2 Pro 10-core CPU and 32 GB of RAM in R 4.4.0 and the default \pkg{BLAS} library packaged with \R \citep{r2024}.
